The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-20 Independence_CH/main_results.tex
Updating counts (wrt no_datatype). Yet another way of presenting instances

\subsection{32 replacement instances to rule them all}
\label{sec:repl-instances}

We isolated 32 instances of Replacement that are sufficient to force
$\CH$ or $\neg\CH$, which are enumerated below by the name of the
corresponding internalized first order formula. Many of these were already present in
relational form in the \session{ZF-Constructible} library.

The first 13 instances are needed to set up
cardinal arithmetic in $M$, along some other basic constructions.

\begin{itemize}
\item 6 instances for iteration of operations (between parentheses). For each construct,
  Paulson used one replacement instance in order to have absoluteness,
  and a second one to obtain closure.
  \begin{itemize}
  \item
    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm} and 
    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.

    For lists on a set $A$ ($X\mapsto \{0 \} \oplus A \times X$, where
    $D\oplus E$ stands for the disjoint union/coproduct $(\{0 \}\times
    D) \cup (\{1 \}\times E)$).
    %
  \item
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm} and
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.

    For (codes of) first order formulas ($X\mapsto (\omega \times
    \omega \oplus \omega \times \omega) \oplus X \times X \oplus X$).
    %
  \item
    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm} and 
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.

    For transitive closure ($X\mapsto\union X$).
  \end{itemize}
\item \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}.

  Absoluteness of the definition of the
  $n$th element of a list (iteration of the tail operation).
  By transitivity, we do not need an instance for closure.
\end{itemize}
%% The instances so far
%% are needed to interpret locales \locale{M{\uscore}datatypes} and
%% \locale{M{\uscore}eclose}.
%% \locale{M{\uscore}datatypes} is used in the relative definition of
%% \isatt{Vset}.
These formulas, together with the next two, form the set
\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms}.
\begin{itemize}
\item
  4 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item \isatt{wfrec{\uscore}rank{\uscore}fm}.

@@

    For the cumulative hierarchy (rank initial segments).
    %
  \item \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.

    For ordertypes
    %
  \item
    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.

    For Aleph.
  \end{itemize}
\item
  2 instances for ordertypes.
  \begin{itemize}
  \item
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.

    Auxiliary instance for the definition of ordertypes, needed for
    the interpretation of \locale{M{\uscore}ordertype}.
    %
  \item
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.

    Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
  \end{itemize}
\end{itemize}

We also need a one extra replacement instance $\psi$ on $M$ for each
$\phi$ of the
previous ones to have them in $M[G]$.
\[
  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
\]
Here, $\mathit{fst}(\lb a,b\rb) = a$, $\mathit{snd}(\lb a,b\rb) = b$
(with default value $0$ for non pairs).
In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
instances appear in the locale \locale{M{\uscore}ZF3} and form the set
\isa{instances{\isadigit{3}}{\isacharunderscore}{\kern0pt}fms}. These are expressed using
the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
obtained by replacing $\phi$ by
$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).

That makes 26 instances up to now. For the setup of forcing, we
require the following 3 instances, which form the set \isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
\begin{itemize}
\item 2 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.

    For check-names.
    %
  \item \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.

    Definition of forcing for atomic formulas.
  \end{itemize}
  %
\item
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.

  Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
  definition of $\punto{G}$).
  %
\end{itemize}
The next two formulas, which form the set
\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}notCH{\isacharunderscore}{\kern0pt}fms},
are needed for the $\Delta$-System Lemma.
\begin{itemize}
\item
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.

@@

  Absoluteness of the previous construction.
\end{itemize}
%
The $31$ formulas up to this point are collected into the set
\isa{overhead{\isacharunderscore}{\kern0pt}notCH}, which is enough to
force $\neg\CH$. To force $\CH$, we required one further instance:
%
\begin{itemize}
\item
  \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.

@@

The particular choice of some of the instances above arose from
Paulson's architecture on which we based our development.
This applies every time
a locale from \session{ZF-Constructible} has to be
interpreted (namely \locale{M{\uscore}datatypes}
\locale{M{\uscore}eclose}, and \locale{M{\uscore}ordertype}).
%% For instance, the first
%% instance required for the definition of relative ordertypes arises
%% from Paulson's \session{ZF-Constructible}.
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139

On the other hand, as explained in
Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
instance arising from the \locale{M{\uscore}basic} locale. Similarly,
we replaced the original proof of the Schröder-Bernstein Theorem by
Zermelo's one \cite[Exr. x4.27]{moschovakis1994notes}, because the
former required at least one extra instance
% (\isatt{banach{\uscore}iterates{\uscore}fm})
arising from an iteration.

It is to be noted that application of the Forcing Theorems do not
require any extra replacement instances on $M$ (independently of the
formula $\phi$ for which they are invoked). This is not the case for
Separation, at least from our formalization: More instances are needed
as the complexity of $\phi$ grows. One point where this is apparent is
in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
that appears as the \isatt{truth{\uscore}lemma} in our development; it
depends on \isatt{truth{\uscore}lemma'} and
\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
\isatt{separation{\uscore}ax}.

@@

Here, the 12-element set \isatt{overhead} is enough to construct a proper
extension. It is  the union of
\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms} and
\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
Also,
\isatt{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
next theorem, the relevant set of formulas is
\isatt{overhead{\isacharunderscore}{\kern0pt}notCH}, defined above in
Section~\ref{sec:repl-instances}, and \isatt{ZC} denotes Zermelo set
theory plus Choice:

>>>>

\subsection{22 replacement instances to rule them all}
\label{sec:repl-instances}

We isolated 22 instances of Replacement that are sufficient to force
$\CH$ or $\neg\CH$, which are enumerated below by the name of the
corresponding internalized first order formula. Many of these were already present in
relational form in the \session{ZF-Constructible} library.

The first 4 instances, collected in the subset
\isatt{instances1{\uscore}fms} of \formula, consist of basic
constructions.

\begin{itemize}
\item 2 instances for transitive closure.
  \begin{itemize}
  \item
    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm}.

    To prove closure under iteration of $X\mapsto\union X$.
  \item
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.

    Auxiliary instance used to show absoluteness.
  \end{itemize}
%% The instances so far
%% are needed to interpret locale
%% \locale{M{\uscore}eclose}.
\item \isatt{wfrec{\uscore}rank{\uscore}fm}.

@@

  For the cumulative hierarchy (rank initial segments).
\end{itemize}

%% The last two and next pair have the same syntactic structure, because
%% they are definitions by well-founded recursion.
The next 4 instances (gathered in \isatt{instances2{\uscore}fms})
are needed to set up
cardinal arithmetic in $M$.
\begin{itemize}
\item 2 instances for ordertypes:
  \begin{itemize}
  \item
    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.

    Well-founded recursion for the construction of ordertypes.
    %
  \item
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.

    Auxiliary instance for the definition of ordertypes.
  \end{itemize}
\item 2 instances for Aleph:
  \begin{itemize}
    %
  \item
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.

    Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem.
    %
  \item
    \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.

    The well-founded recursion to define Aleph.
  \end{itemize}
\end{itemize}

We also need a one extra replacement instance $\psi$ on $M$ for each
$\phi$ of the
previous ones to have them in $M[G]$.
\[
  \psi(x,\alpha,y_1,\dots,y_n) \defi \quine{\alpha = \min \bigl\{
    \beta \mid \exists\tau\in V_\beta.\  \mathit{snd}(x) \forces
    \phi\ [\mathit{fst}(x),\tau,y_1,\dots,y_n]\bigr\}}
\]
Here, $\mathit{fst}(\lb a,b\rb) = a$, $\mathit{snd}(\lb a,b\rb) = b$
(with default value $0$ for non pairs).
In our development, the mapping $\phi\mapsto\psi$ defined above is given by the
$\isatt{ground{\uscore}repl{\uscore}fm}$ function, and all ground replacement
instances appear in the locale \locale{M{\uscore}ZF3} and form the set
\isatt{instances3{\uscore}fms}. These are expressed using
the \isatt{ground{\uscore}replacement{\uscore}assm(M,\isasymphi)} predicate
obtained by replacing $\phi$ by
$\isatt{ground{\uscore}repl{\uscore}fm}(\phi)$ in Eq.~(\ref{eq:replacement_assm_def}).

That makes 16 instances up to now. For the setup of forcing, we
require the following 3 instances, which form the set
\isatt{instances{\uscore}ground{\uscore}fms}.

\begin{itemize}
\item \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.

  Well-founded recursion to define check-names.
  %
\item \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.

  Well-founded recursion for the definition of forcing for atomic formulas.
  %
\item
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.

  Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
  definition of $\punto{G}$).
  %
\end{itemize}
The next two formulas, which form the set
\isatt{instances{\uscore}ground{\uscore}notCH{\uscore}fms},
are needed for the $\Delta$-System Lemma.
\begin{itemize}
\item
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.

@@

  Absoluteness of the previous construction.
\end{itemize}
%
The $21$ formulas up to this point are collected into the set
\isatt{overhead{\uscore}notCH}, which is enough to
force $\neg\CH$. To force $\CH$, we required one further instance:
%
\begin{itemize}
\item
  \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.

@@

The particular choice of some of the instances above arose from
Paulson's architecture on which we based our development.
This applies every time
a locale from \session{ZF-Constructible} has to be
interpreted (\locale{M{\uscore}eclose} and
\locale{M{\uscore}ordertype}, respectively, for the “auxiliary” instances).
%% For instance, the first
%% instance required for the definition of relative ordertypes arises
%% from Paulson's \session{ZF-Constructible}.
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139

On the other hand, as explained in
Section~\ref{sec:zfc-axioms-as-locales}, we managed to eliminate the
instance arising from the \locale{M{\uscore}basic} locale. Similarly,
we replaced the original proof of the Schröder-Bernstein Theorem by
Zermelo's one \cite[Exr. x4.27]{moschovakis1994notes}, because the
former required at least one extra instance
% (\isatt{banach{\uscore}iterates{\uscore}fm})
arising from an iteration. We also managed to avoid 10 further
replacements by restructuring some of original theories in
\session{ZF-Constructible}, so these modifications are included as
part of our project.

It is to be noted that the application of the Forcing Theorems do not
require any extra replacement instances on $M$ (independently of the
formula $\phi$ for which they are invoked). This is not the case for
Separation, at least from our formalization: More instances are needed
as the complexity of $\phi$ grows. One point where this is apparent is
in the proof of Theorem~\ref{th:forcing-thms}(\ref{item:truth-lemma}),
that appears as the \isatt{truth{\uscore}lemma} in our development; it
depends on \isatt{truth{\uscore}lemma'} and
\isatt{truth{\uscore}lemma{\uscore}Neg} which explicitly invoke
\isatt{separation{\uscore}ax}.

@@

Here, the 7-element set \isatt{overhead} is enough to construct a proper
extension. It is  the union of
\isa{instances{\isadigit{1}}{\isacharunderscore}{\kern0pt}fms} and
\isa{instances{\isacharunderscore}{\kern0pt}ground{\isacharunderscore}{\kern0pt}fms}.
Also,
\isatt{{\isasymcdot}Z{\isasymcdot}} denotes Zermelo set theory. In the
next theorem, the relevant set of formulas is
\isatt{overhead{\isacharunderscore}{\kern0pt}notCH}, defined above in
Section~\ref{sec:repl-instances}, and \isatt{ZC} denotes Zermelo set
theory plus Choice:
